Identifying Extracellular Vesicles of Placenta and Uterine Origin Using RNA Expression Signatures

### 2015-09-25 slides/slides.Rnw
finalize slides

\title{Identifying Extracellular Vesicles of Placenta and Uterine 
  Origin Using RNA Expression Signatures}
\author{Don L. Armstrong, Monica Uddin, Derek Wildman}
\institute{Institute for Genomic Biology, Computing Genomes 
  for Reproductive Health, University of Illinois, Urbana-Champaign}
\subject{RNA expression}

@@

\begin{frame}{Identifying source tissue of \acp{EV}}

  One requirement for remotely sensing the transcriptome is
  identifying the source tissue of a specific \ac{EV}. To identify the
  source tissue of a specific \ac{EV}, we have identified miRNA and
  mRNA markers which are characteristic of multiple organs and tissues
  from existing publicly-available RNAseq transcriptomes by
  calculating a tissue specificity index.

  The tissue specificity index for gene $g$ ($\tau_g$) is

  \begin{equation}
    \tau_g = \frac{\sum_{i=1}^{i=N} \left(1-\frac{X_{i,g}}{\mathrm{max}(X_g)}\right)}{N-1}
  \end{equation}
  \label{eq:tissue_specificity_index}

  where $N$ is the number of tissues, $X_{i,g}$ is the mean expression
  of gene $g$ in tissue $i$ and $\mathrm{max}(X_g)$ is the maximal
  expression of gene $g$ across all tissues.

@@

 \begin{frame}{\ac{SVM}}
    We also trained a multiclass \ac{SVM} to accurately identify the
    tissue of origin, and present the results of the identification of
    reads from uterine and placental origin here. Tables showing
    tissue specific genes from all %
    %
    \Sexpr{length(unique(na.omit(tissue.specific.genes[,tissue])))} %
    % 
    tissues can be found on \url{http://dla2.us/p/em2015}.

%     Our approach identified MIR1914, MIR126, SNORD12, and SNORD5 as
%     uterus specific. CGB, LGALS13, and members of the PSG family were
%     identified as placenta specific. The accuracy of the classifier
%     was verified using RNAseq transcriptome data not in the training
%     set. We show the applicability of the classifier to simulated EVs
%     by down sampling reads to $10^5$ reads and biasing the reads to
%     account for preferential association of RNA motifs with EV
%     packaging machinery. When coupled with single-EV sequencing, our
%     classifier enables clinicians to monitor disease-specific markers
%     over the course of treatment without biopsies, increasing the
%     ability of clinicians to adapt treatment modalities to the
%     individual patient. 
\end{frame}

@@

  \item Samples from Roadmap Epigenomics
    \cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping}, GTEx
    \cite{Gtexconsortium.2015:Thegenotypetissueexpression},
    uterine (GSE50599,
    \cite{Chan.ea2014:Assessmentofmyometrialtranscriptome}) and
    placenta (GSE66622,
    \cite{Hughes.ea2015:Evaluatingintra-andinter-individual})
  \item aligned to GRCh38 using STAR
    \cite{Dobin.ea2013:STARultrafastuniversalRNA-seq}
  \item called using Cufflinks 2 v2.0.10
    \cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.

@@

\begin{frame}{Generating SVM}
  The \ac{SVM} was generated using caret
    \cite{Kuhn2008:Buildingpredictivemodelsinr} and svmRadial
    \cite{Karatzoglou2004:kernlab} with 10 cycles of 10 repeats and a
    tune length of 8 with a training set representing $\approx 75\%$ of the
    samples ($\approx 2450$ samples total, including $\approx 35$
    uterine or $\approx 60$ placenta) in each of the groups chosen at
    random, with the remaining samples making up a testing set
    ($\approx 820$ samples).
  \end{frame}

@@

\begin{frame}{Reducing Reads}
\end{frame}

\begin{frame}{Identification of tissue in reduced reads}
  \centering
<<placenta.roc.reduced.curve,fig.width=8,fig.height=4,out.height="0.6\\textheight",echo=FALSE>>=
layout(matrix(1:2,1,2))
for (a in c("placenta","uterus")) {
pl.roc.obj1 <- 
    plot.roc(as.numeric(predictions.caret.svm.reduced[[a]][["testing.1000.reads"]][,"actual"])-1,
             as.numeric(predictions.caret.svm.reduced[[a]][["testing.1000.reads"]][,a]),
             percent=TRUE,
             ## main="Training vs Testing set for Uterus",
             col="#002058"
             )
pl.roc.obj2 <- 
    lines.roc(as.numeric(predictions.caret.svm.reduced[[a]][["testing.5000.reads"]][,"actual"])-1,
              as.numeric(predictions.caret.svm.reduced[[a]][["testing.5000.reads"]][,a]),
              percent=TRUE,
              col="#E87722")
pl.roc.obj3 <- 
    lines.roc(as.numeric(predictions.caret.svm.reduced[[a]][["testing.10.reads"]][,"actual"])-1,
              as.numeric(predictions.caret.svm.reduced[[a]][["testing.10.reads"]][,a]),
              percent=TRUE,
              col="green")
pl.roc.obj4 <- 
    lines.roc(as.numeric(predictions.caret.svm.reduced[["placenta"]][["testing.1.reads"]][,"actual"])-1,
              as.numeric(predictions.caret.svm.reduced[["placenta"]][["testing.1.reads"]][,"placenta"]),
              percent=TRUE,
              col="brown")
# testobj <- roc.test(rocobj1, rocobj2)
# text(50, 50, labels=paste("p-value =", format.pval(testobj$p.value)), adj=c(0, .5))

legend("bottomright",       legend=paste0(capwords(a)," ",c("1 Read","10 Reads","1000 Reads", "5000 Reads")),
       col=c("brown","green","#002058", "#E87722"),
       lwd=2)
}
@ 
\captionof{figure}{\ac{ROC} curve of reduced reads in Placenta and
  Uterus samples with the indicated number of total reads. Reads were
  reduced by sampling reads for each sample weighted by the FPKM
  measured for each sample, and run using the same trained SVM
  depicted in Figure \ref{fig:plac_ut_full_roc}. %
  % 
  The placenta \ac{SVM} is fairly accurate at distinguishing between
  samples of placental origin and non-placental origin even when the
  number of reads is very low (10). The uterus \ac{SVM} requires many
  more reads in order to achieve similar accuracy. This is likely due
  to the much smaller number of uterus specific genes in comparison to
  placenta.}

@@

\section*{Acknowledgements}

\begin{frame}
  The code and underlying data for this poster can be found at
\begin{center}
\qrcode[padding]{http://dla2.us/p/em2015}
\url{http://dla2.us/p/em2015}
\end{center}
\end{frame}

@@

\begin{frame}{Acronyms}
  \acrodef{EV}[EV]{Extracellular Vesicle}
\acrodef{SVM}{Support Vector Machine}
\acrodef{ROC}{Receiver Operating Characteristic}
\acrodef{KNN}{K Nearest Neighbors}
\acrodef{SRA}{Sequence Read Archive}

>>>>

\title[EVs of Placenta and Uterus]{Identifying Extracellular Vesicles of Placenta and Uterine 
  Origin Using RNA Expression Signatures}
\author[Don Armstrong et. al.]{Don L. Armstrong, Monica Uddin, Derek Wildman}
\institute[IGB]{Institute for Genomic Biology, Computing Genomes 
  for Reproductive Health, University of Illinois, Urbana-Champaign}
\subject{RNA expression}

@@

\begin{frame}{Identifying source tissue of \acp{EV}}
  \begin{enumerate}
  \item We want to remotely sense the transcriptome of tissues
  \item For each \ac{EV}, identify the source tissue

  \item So, identify miRNA and mRNA markers which are characteristic of
    multiple organs and tissues
    \begin{itemize}    
    \item Use publicly-available RNAseq transcriptomes

    \item Identify characteristic markers by calculating a tissue
      specificity index for each gene $g$ ($\tau_g$):

      \begin{equation}
        \tau_g = \frac{\sum_{i=1}^{i=N} \left(1-\frac{X_{i,g}}{\mathrm{max}(X_g)}\right)}{N-1}
      \end{equation}
      \label{eq:tissue_specificity_index}
 \item $N$ is the number of tissues, $X_{i,g}$ is the mean expression
   of gene $g$ in tissue $i$ and $\mathrm{max}(X_g)$ is the maximal
   expression of gene $g$
    \end{itemize}
  \end{enumerate}

@@

\begin{frame}{Source of \acp{EV} --- Machine Learning}
  \begin{itemize}
  \item High tissue specificity might not be the best discriminator
  \item Use machine learning techniques!
  \item Train a \ac{SVM}
  \item Use genes which are highly expressed and have either
    \begin{itemize}
    \item High Tissue Specific Index ($\tau_g$)
    \item High Entropy
    \end{itemize}
  \end{itemize}
\end{frame}

@@

  \item Samples from Roadmap Epigenomics
    \cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping}, GTEx
    \cite{Gtexconsortium.2015:Thegenotypetissueexpression},
    uterine (GSE50599,
    \cite{Chan.ea2014:Assessmentofmyometrialtranscriptome}) and
    placenta (GSE66622,
    \cite{Hughes.ea2015:Evaluatingintra-andinter-individual})
  \item aligned to GRCh38 using STAR
    \cite{Dobin.ea2013:STARultrafastuniversalRNA-seq}
  \item quantified using Cufflinks 2 v2.0.10
    \cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.

@@

\begin{frame}{Generating SVM}
 \begin{itemize}
 \item \ac{SVM} was generated using caret
    \cite{Kuhn2008:Buildingpredictivemodelsinr} and svmRadial
    \cite{Karatzoglou2004:kernlab}
  \item 10 cycles of 10 repeats and a
    tune length of 8

  \item training set representing $\approx 75\%$ of the samples
    ($\approx 2450$ samples total, including $\approx 35$ uterine or
    $\approx 60$ placenta) in each of the groups chosen at random,
  \item remaining samples making up a testing set ($\approx 820$
    samples)
  \end{itemize}
\end{frame}

@@

\begin{frame}{Reducing Reads}
  \begin{itemize}
  \item Reads sub-sampled from each sample 
  \item Weighted by the FPKM measured
  \item Run using the same trained SVM shown previously
  \end{itemize}
\end{frame}

\begin{frame}{Identification of tissue in reduced reads}

\begin{columns}
\column{0.6\textwidth}
<<placenta.roc.reduced.curve,fig.width=6,fig.height=6,out.height="0.8\\textheight",echo=FALSE>>=
#layout(matrix(1:2,1,2))
for (a in c("placenta")) {
pl.roc.obj1 <- 
    plot.roc(as.numeric(predictions.caret.svm.reduced[[a]][["testing.1000.reads"]][,"actual"])-1,
             as.numeric(predictions.caret.svm.reduced[[a]][["testing.1000.reads"]][,a]),
             percent=TRUE,
             ## main="Training vs Testing set for Uterus",
             col="#002058"
             )
pl.roc.obj2 <- 
    lines.roc(as.numeric(predictions.caret.svm.reduced[[a]][["testing.5000.reads"]][,"actual"])-1,
              as.numeric(predictions.caret.svm.reduced[[a]][["testing.5000.reads"]][,a]),
              percent=TRUE,
              col="#E87722")
pl.roc.obj3 <- 
    lines.roc(as.numeric(predictions.caret.svm.reduced[[a]][["testing.10.reads"]][,"actual"])-1,
              as.numeric(predictions.caret.svm.reduced[[a]][["testing.10.reads"]][,a]),
              percent=TRUE,
              col="green")
pl.roc.obj4 <- 
    lines.roc(as.numeric(predictions.caret.svm.reduced[["placenta"]][["testing.1.reads"]][,"actual"])-1,
              as.numeric(predictions.caret.svm.reduced[["placenta"]][["testing.1.reads"]][,"placenta"]),
              percent=TRUE,
              col="brown")
# testobj <- roc.test(rocobj1, rocobj2)
# text(50, 50, labels=paste("p-value =", format.pval(testobj$p.value)), adj=c(0, .5))

legend("bottomright",       legend=paste0(capwords(a)," ",c("1 Read","10 Reads","1000 Reads", "5000 Reads")),
       col=c("brown","green","#002058", "#E87722"),
       lwd=2)
}
@ 
\column{0.4\textwidth}
\captionof{figure}{\ac{ROC} curve of reduced reads in Placenta with the indicated number of total reads.  %
  % 
  The placenta \ac{SVM} is fairly accurate at distinguishing between
  samples of placental origin and non-placental origin even when the
  number of reads is very low (10).}
\end{columns}
\end{frame}

\begin{frame}{Reduced Reads: Uterus}
\begin{columns}
\column{0.6\textwidth}
<<uterus.roc.reduced.curve,fig.width=6,fig.height=6,out.height="0.8\\textheight",echo=FALSE>>=
# layout(matrix(1:2,1,2))
for (a in c("uterus")) {
pl.roc.obj1 <- 
    plot.roc(as.numeric(predictions.caret.svm.reduced[[a]][["testing.1000.reads"]][,"actual"])-1,
             as.numeric(predictions.caret.svm.reduced[[a]][["testing.1000.reads"]][,a]),
             percent=TRUE,
             ## main="Training vs Testing set for Uterus",
             col="#002058"
             )
pl.roc.obj2 <- 
    lines.roc(as.numeric(predictions.caret.svm.reduced[[a]][["testing.5000.reads"]][,"actual"])-1,
              as.numeric(predictions.caret.svm.reduced[[a]][["testing.5000.reads"]][,a]),
              percent=TRUE,
              col="#E87722")
pl.roc.obj3 <- 
    lines.roc(as.numeric(predictions.caret.svm.reduced[[a]][["testing.10.reads"]][,"actual"])-1,
              as.numeric(predictions.caret.svm.reduced[[a]][["testing.10.reads"]][,a]),
              percent=TRUE,
              col="green")
pl.roc.obj4 <- 
    lines.roc(as.numeric(predictions.caret.svm.reduced[["placenta"]][["testing.1.reads"]][,"actual"])-1,
              as.numeric(predictions.caret.svm.reduced[["placenta"]][["testing.1.reads"]][,"placenta"]),
              percent=TRUE,
              col="brown")
# testobj <- roc.test(rocobj1, rocobj2)
# text(50, 50, labels=paste("p-value =", format.pval(testobj$p.value)), adj=c(0, .5))

legend("bottomright",       legend=paste0(capwords(a)," ",c("1 Read","10 Reads","1000 Reads", "5000 Reads")),
       col=c("brown","green","#002058", "#E87722"),
       lwd=2)
}
@ 
\column{0.4\textwidth}
\captionof{figure}{\ac{ROC} curve of reduced reads in Placenta and
  Uterus samples with the indicated number of total reads.  %
  % 
  The uterus \ac{SVM} requires many more reads in order to achieve
  similar accuracy. This is likely due to the much smaller number of
  uterus specific genes in comparison to placenta.}
\end{columns}
\end{frame}

@@

\section*{Acknowledgments}

\begin{frame}{Acknowledgments}
  \begin{itemize}
  \item Derek Wildman
  \item Monica Uddin
  \end{itemize}

  The code and underlying data for this talk can be found at
  \begin{center}
    \qrcode[padding]{http://dla2.us/p/em2015}
    \url{http://dla2.us/p/em2015}
  \end{center}
\end{frame}

@@

\begin{frame}[plain]
  \acrodef{EV}[EV]{Extracellular Vesicle}
\acrodef{SVM}{Support Vector Machine}
\acrodef{ROC}{Receiver Operating Characteristic}
\acrodef{KNN}{K Nearest Neighbors}
\acrodef{SRA}{Sequence Read Archive}
